Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-12 inc/ms_model/ms_model.tex
massive spellchecking

This chapter introduces the Analytic Gaussian (AG) model of Michalik and Sipe~\cite{michalik_analytic_2006}.
In the AG model the dynamics of a Gaussian electon pulse are computed using a mean-field self-similar approximation.
The result is a set of equations that can be solved very quickly using numerical methods on a modern computer.
When checked against N-body simulations, the authors find that the model is also surprisingly accurate, even for a large set of non-Gaussian pulses~\cite{michalik_analytic_2006,michalik_evolution_2009}.

@@

For a classical set of $N$ discreet particles (labelled by the index $i$), distributed in 6D space (3 position, 3 momentum) the distribution function is 
\begin{equation}
  f_{\smallD}(\vec{r}, \vec{p}; t) = 
  \sum\limits^{N}_{i=1} \delta(\vec{r}_{i} - \vec{r}) \delta(\vec{p}_{i} - \vec{p}) \,\text{,}
\end{equation}
which automatically satisfies \ref{eq:normalization}.
If the particles have mass $m$ which interact with some potential $V(\vec{r}_{i} - \vec{r}_{j})$, their motion is governed by the equations
\begin{subequations}\label{eq:std_motion}
\begin{gather}
  \frac{d\vec{r}_{i}(t)}{dt} = \frac{\vec{p}_{i}(t)}{m}\\
  \frac{d\vec{p}_{i}(t)}{dt} = -\sum\limits_{j} \frac{d}{d\vec{r}_{i}} V(\vec{r}_{i} - \vec{r}_{j}) \label{eq:std_motion_v} \,\text{.}
\end{gather}
\end{subequations}
Now let us consider the time evolution of $f_{\smallD}$;
\begin{equation}
  \frac{\partial f_{\smallD}}{\partial t} = 
  \frac{d\vec{r}}{dt} \frac{\partial f_{\smallD}}{\partial \vec{r}} 
  + \frac{\partial f_{\smallD}}{\partial \vec{p}} \frac{d\vec{p}}{dt} \,\text{,}
\end{equation}
but with \ref{eq:std_motion} the final discreet evolution equation is
\begin{equation} \label{eq:evolution_discreet}
  \frac{d}{dt} f_{\smallD}(\vec{r}, \vec{p}; t) =
  -\frac{\vec{p}}{m} \frac{\partial}{\partial \vec{r}} f_{\smallD}(\vec{r}, \vec{p}; t)
  + \frac{\partial}{\partial \vec{p}} f_{\smallD} (\vec{r}, \vec{p}; t)
  \frac{\partial}{\partial \vec{r}} \iint d\vecprime{r} d\vecprime{p} V(\vec{r} - \vecprime{r}) f_{\smallD}(\vecprime{r},\vecprime{p};t) \,\text{,}
\end{equation}
where the extra negation comes from the change of perspective of \ref{eq:std_motion_v} from the frame of the individual particle $i$ to the observer frame.

This analysis can be extended to continuous distributions by taking the ensemble average over \ref{eq:evolution_discreet}.
In order to simplify the resulting equation to a useful result, one applies the usual mean-field approximation
\begin{equation}
  \left < f_{\smallD} (\vec{r}, \vec{p}; t) f_{\smallD} (\vecprime{r}, \vecprime{p}; t) \right > = \left < f_{\smallD} (\vec{r}, \vec{p}; t) \right > \left < f_{\smallD} (\vecprime{r}, \vecprime{p}; t) \right > \,\text{,}
\end{equation}
so that the formal evolution of a continuous distribution is given by
\begin{equation} \label{eq:formal_evolution}
  \frac{d}{dt} f(\vec{r}, \vec{p}; t) =
  -\frac{\vec{p}}{m} \frac{\partial}{\partial \vec{r}} f(\vec{r}, \vec{p}; t)
  + \frac{\partial}{\partial \vec{p}} f (\vec{r}, \vec{p}; t)
  \frac{\partial}{\partial \vec{r}} \iint d\vecprime{r} d\vecprime{p} V(\vec{r} - \vecprime{r}) f(\vecprime{r},\vecprime{p};t) \,\text{.}
\end{equation}
Although this equation looks very similar to \ref{eq:evolution_discreet}, notice that it contains approximations, mean-field that has been seen and in practice the self-similarity of the distribution function, that the discreet discription did not.

@@

The formal evolution explored in Section \ref{sec:formal_evolution} allows for any distribution function.
For Ultrafast Electron Microscopy, the electron pulse is generated by a laser pulse, therefore the electron pulse can be thought to be a Gaussian in space and time, mimicing the laser pulse which generated it.
How correct this assumption is will be explored in \ref{sec:initial_conditions}.
In this section we consider a Gaussian distribution function, generally of the form
\begin{equation}
  f(\vec{r}, \vec{p}; t) = C(t)\exp \left[ - G(\vec{r}, \vec{p}; t) \right] \,\text{.}
\end{equation}
The analysis presented here will consider a cylindrically symmetric pulse.
For the remainder of this analysis, the subscripts $\smallT$ (transverse) and $z$ (longitudinal) will accompany many quantites, reflecting the congruence of $x$ and $y$.
The subscript $\alpha$ will refer to either $\smallT$ or $z$.
In order to model several important physical characteristics of the pulse, we choose the argument (i.e. the function $G$) as
\begin{equation} \label{eq:gaussian_exponent}
\begin{split}
  G(\vec{r}, \vec{p}; t) &= \frac{x^2 + y^2}{2 \sigma_{\smallT}} + \frac{z^2}{2 \sigma_{z}} \\
  &+\frac{
        [p_x - (\gamma_{\smallT}/\sigma_{\smallT}) x ]^2 
      + [p_y - (\gamma_{\smallT}/\sigma_{\smallT}) y ]^2
    }{2 \eta_{\smallT}}
  + \frac{ [p_z - (\gamma_{z}/\sigma_{z}) z ]^2 }{2 \eta_{z}} \,\text{.}
\end{split}
\end{equation}
This parameterization defines the spatial variance $\sigma_{\alpha}$, the local momentum variance $\eta_{\alpha}$, and the momentum chirp $\gamma_{\alpha}$ in each spatial direction $\alpha$.
The ``chirp'' of a pulse quantifies the difference in average local momentum from one side of the pulse to the other.
For example, in a pulse which is normally (positively) chirped longitudinally, electrons near the front of the pulse will be moving faster (on average) than those in the back.
The normalization constant $C(t)$, for this distributions of $N$ total particles, is found by applying equation \ref{eq:normalization}:
\begin{equation}
  C(t) = \frac{N}{(2\pi)^3} 
  \left( 
    \frac{1}{\sigma_{\smallT}^2\eta_{\smallT}^2\sigma_{z}\eta_{z}}
  \right)^{\frac{1}{2}} \text{.}
\end{equation}

>>>>

This chapter introduces the Analytic Gaussian (AG) model of Michalik and Sipe~\cite{michalik_analytic_2006}.
In the AG model the dynamics of a Gaussian electron pulse are computed using a mean-field self-similar approximation.
The result is a set of equations that can be solved very quickly using numerical methods on a modern computer.
When checked against N-body simulations, the authors find that the model is also surprisingly accurate, even for a large set of non-Gaussian pulses~\cite{michalik_analytic_2006,michalik_evolution_2009}.

@@

For a classical set of $N$ discreet particles (labeled by the index $i$), distributed in 6D space (3 position, 3 momentum) the distribution function is 
\begin{equation}
  f_{\smallD}(\vec{r}, \vec{p}; t) = 
  \sum\limits^{N}_{i=1} \delta(\vec{r}_{i} - \vec{r}) \delta(\vec{p}_{i} - \vec{p}) \,\text{,}
\end{equation}
which automatically satisfies \ref{eq:normalization}.
If the particles have mass $m$ which interact with some potential $V(\vec{r}_{i} - \vec{r}_{j})$, their motion is governed by the equations
\begin{subequations}\label{eq:std_motion}
\begin{gather}
  \frac{d\vec{r}_{i}(t)}{dt} = \frac{\vec{p}_{i}(t)}{m}\\
  \frac{d\vec{p}_{i}(t)}{dt} = -\sum\limits_{j} \frac{d}{d\vec{r}_{i}} V(\vec{r}_{i} - \vec{r}_{j}) \label{eq:std_motion_v} \,\text{.}
\end{gather}
\end{subequations}
Now let us consider the time evolution of $f_{\smallD}$;
\begin{equation}
  \frac{\partial f_{\smallD}}{\partial t} = 
  \frac{d\vec{r}}{dt} \frac{\partial f_{\smallD}}{\partial \vec{r}} 
  + \frac{\partial f_{\smallD}}{\partial \vec{p}} \frac{d\vec{p}}{dt} \,\text{,}
\end{equation}
but with \ref{eq:std_motion} the final discreet evolution equation is
\begin{equation} \label{eq:evolution_discreet}
  \frac{d}{dt} f_{\smallD}(\vec{r}, \vec{p}; t) =
  -\frac{\vec{p}}{m} \frac{\partial}{\partial \vec{r}} f_{\smallD}(\vec{r}, \vec{p}; t)
  + \frac{\partial}{\partial \vec{p}} f_{\smallD} (\vec{r}, \vec{p}; t)
  \frac{\partial}{\partial \vec{r}} \iint d\vecprime{r} d\vecprime{p} V(\vec{r} - \vecprime{r}) f_{\smallD}(\vecprime{r},\vecprime{p};t) \,\text{,}
\end{equation}
where the extra negation comes from the change of perspective of \ref{eq:std_motion_v} from the frame of the individual particle $i$ to the observer frame.

This analysis can be extended to continuous distributions by taking the ensemble average over \ref{eq:evolution_discreet}.
In order to simplify the resulting equation to a useful result, one applies the usual mean-field approximation
\begin{equation}
  \left < f_{\smallD} (\vec{r}, \vec{p}; t) f_{\smallD} (\vecprime{r}, \vecprime{p}; t) \right > = \left < f_{\smallD} (\vec{r}, \vec{p}; t) \right > \left < f_{\smallD} (\vecprime{r}, \vecprime{p}; t) \right > \,\text{,}
\end{equation}
so that the formal evolution of a continuous distribution is given by
\begin{equation} \label{eq:formal_evolution}
  \frac{d}{dt} f(\vec{r}, \vec{p}; t) =
  -\frac{\vec{p}}{m} \frac{\partial}{\partial \vec{r}} f(\vec{r}, \vec{p}; t)
  + \frac{\partial}{\partial \vec{p}} f (\vec{r}, \vec{p}; t)
  \frac{\partial}{\partial \vec{r}} \iint d\vecprime{r} d\vecprime{p} V(\vec{r} - \vecprime{r}) f(\vecprime{r},\vecprime{p};t) \,\text{.}
\end{equation}
Although this equation looks very similar to \ref{eq:evolution_discreet}, notice that it contains approximations, mean-field that has been seen and in practice the self-similarity of the distribution function, that the discreet description did not.

@@

The formal evolution explored in Section \ref{sec:formal_evolution} allows for any distribution function.
For Ultrafast Electron Microscopy, the electron pulse is generated by a laser pulse, therefore the electron pulse can be thought to be a Gaussian in space and time, mimicking the laser pulse which generated it.
How correct this assumption is will be explored in \ref{sec:initial_conditions}.
In this section we consider a Gaussian distribution function, generally of the form
\begin{equation}
  f(\vec{r}, \vec{p}; t) = C(t)\exp \left[ - G(\vec{r}, \vec{p}; t) \right] \,\text{.}
\end{equation}
The analysis presented here will consider a cylindrically symmetric pulse.
For the remainder of this analysis, the subscripts $\smallT$ (transverse) and $z$ (longitudinal) will accompany many quantities, reflecting the congruence of $x$ and $y$.
The subscript $\alpha$ will refer to either $\smallT$ or $z$.
In order to model several important physical characteristics of the pulse, we choose the argument (i.e. the function $G$) as
\begin{equation} \label{eq:gaussian_exponent}
\begin{split}
  G(\vec{r}, \vec{p}; t) &= \frac{x^2 + y^2}{2 \sigma_{\smallT}} + \frac{z^2}{2 \sigma_{z}} \\
  &+\frac{
        [p_x - (\gamma_{\smallT}/\sigma_{\smallT}) x ]^2 
      + [p_y - (\gamma_{\smallT}/\sigma_{\smallT}) y ]^2
    }{2 \eta_{\smallT}}
  + \frac{ [p_z - (\gamma_{z}/\sigma_{z}) z ]^2 }{2 \eta_{z}} \,\text{.}
\end{split}
\end{equation}
This parameterization defines the spatial variance $\sigma_{\alpha}$, the local momentum variance $\eta_{\alpha}$, and the momentum chirp $\gamma_{\alpha}$ in each spatial direction $\alpha$.
The ``chirp'' of a pulse quantifies the difference in average local momentum from one side of the pulse to the other.
For example, in a pulse which is normally (positively) chirped longitudinally, electrons near the front of the pulse will be moving faster (on average) than those in the back.
The normalization constant $C(t)$, for this distributions of $N$ total particles, is found by applying equation \ref{eq:normalization}:
\begin{equation}
  C(t) = \frac{N}{(2\pi)^3} 
  \left( 
    \frac{1}{\sigma_{\smallT}^2\eta_{\smallT}^2\sigma_{z}\eta_{z}}
  \right)^{\frac{1}{2}} \text{.}
\end{equation}

### 2013-04-12 inc/photocathode/este/este.tex
massive spellchecking

\ref{fig:este-semicond} depicts the results obtained for GaSb and InSb photocathode materials under pulsed UV (261nm) laser irradiation using the same experimental technique employed for PAPE, 2PTE, and single-photon PE studies.
Both samples are cut from [100]-oriented polished wafers; the GaSb is undoped and the InSb is moderately $p$-type.
Prior studies~\cite{gobeli_photoelectric_1965} indicate that single-photon photoemission should not possible for either semiconductor --- the effective photoemission work functions (about 4.8eV for undoped GaSb and 4.89eV for $p$-type InSb (Fermi level pinned at valence band maximum)) being greater than the 4.75eV photon energy even when the 34meV Schottky barrier suppression due to the applied 8kV/cm DC field is included~\cite{dowell_quantum_2009,berger_excited_2012}.
Nonetheless, for both zincblende semiconductors, significant laser-driven emission (more than expected for a single-photon field emission process) is observed with a yield nearly linearly proportional to the $\sim$4ps UV laser pulse energy.
Note that due to the surprisingly large yield, we repeated the measurements, this time replacing the the scintillator screen with a Faraday cup and repeating the laser energy sequence.
In this way, the yields for these photocathodes were measured directly and absolutely, unlike the results presented in the preceeding sections, for which relative measurements were sufficient.

@@

Carrier cooling, primarily by longitudinal optical (LO) phonon emission, and decay out of the upper conduction band to lower states will rapidly deplete the Boltzmann tail above the effective $\sim$1eV work function, thereby curtailing the observed ESTE.
By comparison with the $\Gamma_6$ conduction band Fr\"ohlich coupling constant in GaAs, where the characteristic LO phonon emission time is 165fs for a hot electron in the $\Gamma_6$ condunction band~\cite{kash_subpicosecond_1985}, we estimate that hot electrons in the upper $\Gamma_8$ conduction band of both GaSb and InSb emit a LO phonon every $\sim$200fs with an energy of 29 and 24meV, respectively.
This means that even without population decay mechanisms the temperature of the exited electron distribution $T_e$ drops at a rate of $\sim$1,600K/ps in both semiconductors.
We note that the close proximity of the lower and unconfined (in momentum space) $\Gamma_7$ upper conduction band to the $\Gamma_8$ band will likely result in a fast population decay, so that the observed ESTE process should have an intrinsic latency of much less than the $\sim$4ps UV laser pulse duration.

>>>>

\ref{fig:este-semicond} depicts the results obtained for GaSb and InSb photocathode materials under pulsed UV (261nm) laser irradiation using the same experimental technique employed for PAPE, 2PTE, and single-photon PE studies.
Both samples are cut from [100]-oriented polished wafers; the GaSb is undoped and the InSb is moderately $p$-type.
Prior studies~\cite{gobeli_photoelectric_1965} indicate that single-photon photoemission should not possible for either semiconductor --- the effective photoemission work functions (about 4.8eV for undoped GaSb and 4.89eV for $p$-type InSb (Fermi level pinned at valence band maximum)) being greater than the 4.75eV photon energy even when the 34meV Schottky barrier suppression due to the applied 8kV/cm DC field is included~\cite{dowell_quantum_2009,berger_excited_2012}.
Nonetheless, for both zincblende semiconductors, significant laser-driven emission (more than expected for a single-photon field emission process) is observed with a yield nearly linearly proportional to the $\sim$4ps UV laser pulse energy.
Note that due to the surprisingly large yield, we repeated the measurements, this time replacing the the scintillator screen with a Faraday cup and repeating the laser energy sequence.
In this way, the yields for these photocathodes were measured directly and absolutely, unlike the results presented in the preceding sections, for which relative measurements were sufficient.

@@

Carrier cooling, primarily by longitudinal optical (LO) phonon emission, and decay out of the upper conduction band to lower states will rapidly deplete the Boltzmann tail above the effective $\sim$1eV work function, thereby curtailing the observed ESTE.
By comparison with the $\Gamma_6$ conduction band Fr\"ohlich coupling constant in GaAs, where the characteristic LO phonon emission time is 165fs for a hot electron in the $\Gamma_6$ conduction band~\cite{kash_subpicosecond_1985}, we estimate that hot electrons in the upper $\Gamma_8$ conduction band of both GaSb and InSb emit a LO phonon every $\sim$200fs with an energy of 29 and 24meV, respectively.
This means that even without population decay mechanisms the temperature of the exited electron distribution $T_e$ drops at a rate of $\sim$1,600K/ps in both semiconductors.
We note that the close proximity of the lower and unconfined (in momentum space) $\Gamma_7$ upper conduction band to the $\Gamma_8$ band will likely result in a fast population decay, so that the observed ESTE process should have an intrinsic latency of much less than the $\sim$4ps UV laser pulse duration.

### 2013-04-12 inc/photocathode/plasmon/plasmon.tex
massive spellchecking

This avenue of investigation was motivated by Zawadzka et al~\cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors coat a thin gold film on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the alternative ``grating coupling'' geometry~\cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

At the interface of a vacuum and such a simple material, contstraints on the material parameters which may allow surface plasmons, result in the condition that the dielectric constant at the oscillation frequency of the surface plasmon $ \varepsilon(\omega_{SP}) = -1 $~\cite{cottam_introduction_2004}.
Substituting this result into \ref{eq:dielectric_as_omega} shows that this frequency of oscillation is simply related to intrinsic plasma frequency by
\begin{equation}
  \omega_{SP} = \omega_{P} / \sqrt{2} \,\text{.}
\end{equation}
Further, plasmons propagating along a vacuum interface carry momentum whose magnitude is given by 
\begin{equation}
  k_{SP}^2 = \frac{\omega^2}{c^2} \left( \frac{ \varepsilon }{ 1 + \varepsilon } \right) \,\text{.}
\end{equation}

@@

\subsection{Commerical Grating on Glass}

@@

\begin{figure}
  \centering
  \includegraphics{damage.png}
  \caption{Damage to commerical gold-on-glass holographic grating}
  \label{fig:grating-damage}
\end{figure}

@@

After the destruction of the gold foil on glass grating, new gratings were designed to have a higher subtrate thermal conductivity.
The second generation grating was constructed from a silicon substract then coated with a gold film.
The thermal conductivity of silicon is approximately 100 times that of glass. %TODO reference needed?

@@

\begin{figure}
  \centering
  \input{power_seq}
  \caption[Photoemission characteristics of laser-driven plasmons on a one-dimensional periodic gold surface]{
    Photoemission characteristics of laser-driven plasmons on a one-dimensional periodic gold surface.
    Top: Electron emission strength as a function of incident 523nm femtosecond laser pulse intensity for coherently-driven plasmons (top curve)
      % with parabolic fit to initial dependence (dashed curve)
      and comparative emission from a plane gold surface under the same irradiantion conditions.
    Bottom: Two-dimensional electron momentum distributions for plasmon-assisted emission (A-D) and two-photon field emission from gold (E) at the laser intensities indicated on the top graph.
    The circular aperture visible in B-D is the 6mm-diameter bore of the magnetic focusing lens.
    The orientation of the grating lines of the periodic gold surface with respect to the emission in A-D is parallel to the broadening seen in B-D.
  }
  \label{fig:pape}
\end{figure}

% Content taken from 2010 ANL report
\ref{fig:pape} summarizes the photoemission results obtained from this gold-coated periodic photocathode under green (523nm) laser irradiation.
Strong enhanced phototemission was observed for only a narrow $\pm$5mrad range of incidence angles $\theta$ around the expected plasmon resonance angle of 39$^{\circ}$.
Moreover, this enhanced laser-driven electron emission was only observed for $p$-polarized incident 523nm femtosecond laser radiation.
These two properties are known characteristics of plasmon-assisted photoemission~\cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.

@@

Above an incident laser pulse intensity of $\sim$32MW/cm$^2$, there is a rapid and dramatic increase in the emission efficiency for the nano-patterned photocathode.
This can be explained by a suppression of the barrier to photoemission below the two-photon excitation energy, thus allowing direct two-photon photoemission.
The well-known Schottky Effect states that for an applied electric field $E$ the work function will be suppressed by an amount
\begin{equation}
  \Delta \Phi = - e\sqrt{\frac{e E}{4 \pi \varepsilon_{\smallzero}}} \,\text{.}
\end{equation}
This suppression is likely caused by the laser-excited surface plasmon, since on the neighboring flat gold surface, where the suppression due to the laser (20MV/m) and the 20kV acceleration potential (0.8MV/m) is $\Delta \Phi \approx $ 0.18eV, no similar increase in efficiency is observed.
If this is indeed the case, and the surface plasmon field futher suppresses the barrier to a total $ \Delta \Phi \approx $ 0.35eV to allow direct two-photon photoemission, the field must be of the order of 60MV/m or three times that of the laser field.
In \ref{fig:pape} B, at a laser intensity of $\sim$30MW/cm$^2$, there is already a distinct broadening of the momentum distribution in the direction parallel to the wavevector $k_G$ of the sinusoidal modulation on the photocathode (i.e., perpendicular to the FIB-milled grooves). 
This broadening becomes the dominant feature of the electron emission above $\sim$32MW/cm$^2$; indeed, it becomes so severe that at laser intensities above $\sim$37MW/cm$^2$ the 6mm-diameter bore of the magnetic lenses becomes a strong limiting aperture for the resulting beam --- as is clearly seen in \ref{fig:pape} D and by the cut-off in the detected electron signal (YAG fluorescence).

@@

\begin{figure}
  \centering
  \input{rotated}
  \caption[Comparison of the colinear and rotated plasmonic excitation geometries]{
    Comparison of the collinear (above) and rotated (below) plasmonic excitation geometries.
    This rotation separates the directionality of effects caused by the laser, plasmon field and grating.
    Each vector diagrams (left) is shown next to a representative observed momentum distribution image (right).
    The increased divergence is shown to be rotated by the same $45^{\circ}$ as the grating (the original colinear line is shown dashed); evidence that the grating is the cause of the deleterious effect.
  }
  \label{fig:rotated}
\end{figure}

@@

The experimental results are shown to the right in \ref{fig:rotated}.
As the divergence is clearly shown to be rotated by $45^{\circ}$ with respect to the colinear case, this experiment demonstrates that the cause of the divergence increase is due to the local field enhancements near the peaks of the grating.
For a mathematical treatment of the transverse electric field of such patterned surfaces, the reader is referred to Ref.~\cite{watts_sharp_1997}. 
Our results are commensurate with the results shown by Zawadzka et al~\cite{zawadzka_evanescent_2001}, who showed a reduced divergence in the Kretschmann geometry.
As that geometry does not employ a patterned surface, it would not be subject to such local field enhancements of the accelerating electric field.

@@

In an attempt to mitigate the effect of the surface peaks on the accelerating field, a new pattern was commissioned and prepared in a similar manner to the original.
This new photocathode featured long trenches of an approximately square profile, arranged periodically across the surface; the trench spacing was five times the trench width in an attempt to present a ``flatter'' surface, which would should be less disruptive to the accelerating electric field.
Although this periodicity should have been sufficient to facilitate the generation of the plasmon, the photocathode performed markedly worse than its sinusoidal predecessor.
Indeed, no evidence of sucessful plasmon oscillation was demonstrated after repeated attempts.

>>>>

This avenue of investigation was motivated by Zawadzka et al~\cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors coat a thin gold film on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the alternative ``grating coupling'' geometry~\cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Further, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

At the interface of a vacuum and such a simple material, constraints on the material parameters which may allow surface plasmons, result in the condition that the dielectric constant at the oscillation frequency of the surface plasmon $ \varepsilon(\omega_{SP}) = -1 $~\cite{cottam_introduction_2004}.
Substituting this result into \ref{eq:dielectric_as_omega} shows that this frequency of oscillation is simply related to intrinsic plasma frequency by
\begin{equation}
  \omega_{SP} = \omega_{P} / \sqrt{2} \,\text{.}
\end{equation}
Further, plasmons propagating along a vacuum interface carry momentum whose magnitude is given by 
\begin{equation}
  k_{SP}^2 = \frac{\omega^2}{c^2} \left( \frac{ \varepsilon }{ 1 + \varepsilon } \right) \,\text{.}
\end{equation}

@@

\subsection{Commercial Grating on Glass}

@@

\begin{figure}
  \centering
  \includegraphics{damage.png}
  \caption{Damage to commercial gold-on-glass holographic grating}
  \label{fig:grating-damage}
\end{figure}

@@

After the destruction of the gold foil on glass grating, new gratings were designed to have a higher substrate thermal conductivity.
The second generation grating was constructed from a silicon substrate then coated with a gold film.
The thermal conductivity of silicon is approximately 100 times that of glass. %TODO reference needed?

@@

\begin{figure}
  \centering
  \input{power_seq}
  \caption[Photoemission characteristics of laser-driven plasmons on a one-dimensional periodic gold surface]{
    Photoemission characteristics of laser-driven plasmons on a one-dimensional periodic gold surface.
    Top: Electron emission strength as a function of incident 523nm femtosecond laser pulse intensity for coherently-driven plasmons (top curve)
      % with parabolic fit to initial dependence (dashed curve)
      and comparative emission from a plane gold surface under the same irradiation conditions.
    Bottom: Two-dimensional electron momentum distributions for plasmon-assisted emission (A-D) and two-photon field emission from gold (E) at the laser intensities indicated on the top graph.
    The circular aperture visible in B-D is the 6mm-diameter bore of the magnetic focusing lens.
    The orientation of the grating lines of the periodic gold surface with respect to the emission in A-D is parallel to the broadening seen in B-D.
  }
  \label{fig:pape}
\end{figure}

% Content taken from 2010 ANL report
\ref{fig:pape} summarizes the photoemission results obtained from this gold-coated periodic photocathode under green (523nm) laser irradiation.
Strong enhanced photoemission was observed for only a narrow $\pm$5mrad range of incidence angles $\theta$ around the expected plasmon resonance angle of 39$^{\circ}$.
Moreover, this enhanced laser-driven electron emission was only observed for $p$-polarized incident 523nm femtosecond laser radiation.
These two properties are known characteristics of plasmon-assisted photoemission~\cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.

@@

Above an incident laser pulse intensity of $\sim$32MW/cm$^2$, there is a rapid and dramatic increase in the emission efficiency for the nano-patterned photocathode.
This can be explained by a suppression of the barrier to photoemission below the two-photon excitation energy, thus allowing direct two-photon photoemission.
The well-known Schottky Effect states that for an applied electric field $E$ the work function will be suppressed by an amount
\begin{equation}
  \Delta \Phi = - e\sqrt{\frac{e E}{4 \pi \varepsilon_{\smallzero}}} \,\text{.}
\end{equation}
This suppression is likely caused by the laser-excited surface plasmon, since on the neighboring flat gold surface, where the suppression due to the laser (20MV/m) and the 20kV acceleration potential (0.8MV/m) is $\Delta \Phi \approx $ 0.18eV, no similar increase in efficiency is observed.
If this is indeed the case, and the surface plasmon field further suppresses the barrier to a total $ \Delta \Phi \approx $ 0.35eV to allow direct two-photon photoemission, the field must be of the order of 60MV/m or three times that of the laser field.
In \ref{fig:pape} B, at a laser intensity of $\sim$30MW/cm$^2$, there is already a distinct broadening of the momentum distribution in the direction parallel to the wavevector $k_G$ of the sinusoidal modulation on the photocathode (i.e., perpendicular to the FIB-milled grooves). 
This broadening becomes the dominant feature of the electron emission above $\sim$32MW/cm$^2$; indeed, it becomes so severe that at laser intensities above $\sim$37MW/cm$^2$ the 6mm-diameter bore of the magnetic lenses becomes a strong limiting aperture for the resulting beam --- as is clearly seen in \ref{fig:pape} D and by the cut-off in the detected electron signal (YAG fluorescence).

@@

\begin{figure}
  \centering
  \input{rotated}
  \caption[Comparison of the collinear and rotated plasmonic excitation geometries]{
    Comparison of the collinear (above) and rotated (below) plasmonic excitation geometries.
    This rotation separates the directionality of effects caused by the laser, plasmon field and grating.
    Each vector diagrams (left) is shown next to a representative observed momentum distribution image (right).
    The increased divergence is shown to be rotated by the same $45^{\circ}$ as the grating (the original collinear line is shown dashed); evidence that the grating is the cause of the deleterious effect.
  }
  \label{fig:rotated}
\end{figure}

@@

The experimental results are shown to the right in \ref{fig:rotated}.
As the divergence is clearly shown to be rotated by $45^{\circ}$ with respect to the collinear case, this experiment demonstrates that the cause of the divergence increase is due to the local field enhancements near the peaks of the grating.
For a mathematical treatment of the transverse electric field of such patterned surfaces, the reader is referred to Ref.~\cite{watts_sharp_1997}. 
Our results are commensurate with the results shown by Zawadzka et al~\cite{zawadzka_evanescent_2001}, who showed a reduced divergence in the Kretschmann geometry.
As that geometry does not employ a patterned surface, it would not be subject to such local field enhancements of the accelerating electric field.

@@

In an attempt to mitigate the effect of the surface peaks on the accelerating field, a new pattern was commissioned and prepared in a similar manner to the original.
This new photocathode featured long trenches of an approximately square profile, arranged periodically across the surface; the trench spacing was five times the trench width in an attempt to present a ``flatter'' surface, which would should be less disruptive to the accelerating electric field.
Although this periodicity should have been sufficient to facilitate the generation of the plasmon, the photocathode performed markedly worse than its sinusoidal predecessor.
Indeed, no evidence of successful plasmon oscillation was demonstrated after repeated attempts.

### 2013-04-12 inc/photocathode/setup/setup.tex
massive spellchecking

To address this need, I have experimented with many candidate photocathodes, which implement a variety of different photoemission processes.
In this chapter, I will present an example of several of these different processes, discuss its prospects for reducing the transverse emittance and finally assess its utility as an electron source for UEM.
I will then outline some future research directions that the group at UIC will continue to pursue.
Since this is the most active current field in our group's reseach, some very new results are presented near the end of the chapter and future plans are evolving.

>>>>

To address this need, I have experimented with many candidate photocathodes, which implement a variety of different photoemission processes.
In this chapter, I will present an example of several of these different processes, discuss its prospects for reducing the transverse emittance and finally assess its utility as an electron source for UEM.
I will then outline some future research directions that the group at UIC will continue to pursue.
Since this is the most active current field in our group's research, some very new results are presented near the end of the chapter and future plans are evolving.

### 2013-04-12 inc/photocathode/single_photon/single_photon.tex
massive spellchecking

Single-photon photoemission is the most common photoemission process and the easiest to explain.
As described by Einstein in his famous paper on ``the photoelectric effect''~\cite{einstein_uber_1905}, as it is colloquially known, if an electron absorbs a photon whose energy is sufficient to promote the electron above the local vacuum level, it becomes free to leave the host material.
The material dependent minimum energy needed to eject an electron is termed its ``work function'' $\Phi$.
Any additional energy above the work function remains with the ejected electron, usually in the form of kinetic energy, and is called the ``excess photoemission energy'' $\Delta E$; for single photon photoemission, using a laser of frequency $\omega$ (photon energy $\hbar \omega$), this is simply,
\begin{equation}
  \Delta E = \hbar \omega - \Phi \,\text{.}
\end{equation}
The efficiency of photoemission $\eta_{{\scriptscriptstyle PE}}$, when $\Delta E$ is small compared to $\Phi$ may be written as, %TODO ref
\begin{equation}
  \eta_{{\scriptscriptstyle PE}} = A ( 1 - R ) \Delta E^2 \,\text{,}
\end{equation}
where $A$ is a material paramenter (typically of the order of $\sim 10^{-4} \text{eV}^{-2}$ %TODO ref
), and $R$ is the photocathode surface reflectivity, which is itself a function of the polarization, frequency and angle of the incident light.
%TODO calculation of $\eta_{PE}$?

>>>>

Single-photon photoemission is the most common photoemission process and the easiest to explain.
As described by Einstein in his famous paper on ``the photoelectric effect''~\cite{einstein_uber_1905}, as it is colloquially known, if an electron absorbs a photon whose energy is sufficient to promote the electron above the local vacuum level, it becomes free to leave the host material.
The material dependent minimum energy needed to eject an electron is termed its ``work function'' $\Phi$.
Any additional energy above the work function remains with the ejected electron, usually in the form of kinetic energy, and is called the ``excess photoemission energy'' $\Delta E$; for single photon photoemission, using a laser of frequency $\omega$ (photon energy $\hbar \omega$), this is simply,
\begin{equation}
  \Delta E = \hbar \omega - \Phi \,\text{.}
\end{equation}
The efficiency of photoemission $\eta_{{\scriptscriptstyle PE}}$, when $\Delta E$ is small compared to $\Phi$ may be written as, %TODO ref
\begin{equation}
  \eta_{{\scriptscriptstyle PE}} = A ( 1 - R ) \Delta E^2 \,\text{,}
\end{equation}
where $A$ is a material parameter (typically of the order of $\sim 10^{-4} \text{eV}^{-2}$ %TODO ref
), and $R$ is the photocathode surface reflectivity, which is itself a function of the polarization, frequency and angle of the incident light.
%TODO calculation of $\eta_{PE}$?
